---
title: "Phylodynamics using Serial UPGMA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the FASTA file of your county of interest. 
```{r, echo=FALSE}
library(phylodyn)
library(ape)
library(lubridate)
source("function_serial.R")
#Read fasta file
country<-"fastaFrance"
fastaname<-paste(country,".fasta",sep="")
fastaformat<-read.FASTA(fastaname)
```

Extract sequences, sequence names, and sampling dates from the file
```{r, echo=FALSE}
#Save name of the sequence and sampling times from the file
seq_names<-names(fastaformat)
samp_times<-c()
for (r in seq_names){
  samp_times<-c(samp_times,paste(strsplit(r,"|")[[1]][(length(strsplit(r,"|")[[1]])-9):length(strsplit(r,"|")[[1]])],collapse =""))
}
samp_times<-decimal_date(date(samp_times))
lastdate<-max(samp_times)
samp_times<-max(samp_times)-samp_times
name_samp<-cbind(samp_times,seq_names)
```

Ensure that there are at least two sequences at t=0, otherwise remove the sequence
```{r, echo=FALSE}
while(table(samp_times)[1]==1){
  idx<-which(samp_times==0)
  #Remove the sequence observed only once
  fastaformat <- fastaformat[-idx]
  seq_names<-seq_names[-idx]
  samp_times<-samp_times[-idx]
  name_samp<-name_samp[-idx,]
  #Make the the last sequence the new 0
  samp_times<-samp_times-min(samp_times)
}
```

Compute mutation rate. Note: the reference sequence must be in the last spot
```{r,echo=FALSE}
mu<-mu_linear_reg(fastaformat)
```

Compute serial UPGMA tree and plot it
```{r,echo=FALSE}
fastafile<-as.phyDat(fastaformat)
tree<-serial_upgma(fastafile,mu, samp_times, name_samp, model="F81")
plot(tree,show.tip.label = TRUE,cex=.3)
```

OPTIONAL: Check if there are any branch lengths (terminal ones) from the tips longer than X (cutoff) year. If any, remove them, recompute mu and the tree
```{r,echo=FALSE}
cutoff<-0.35
id.long<-which(tree$edge.length>=cutoff & tree$edge[,2]<=n)
if (length(id.long)>0){
  id.labels<-tree$tip.label[tree$edge[id.long,2]]
  id.remove<-c()
  for (i in id.labels) {id.remove <-  c(id.remove,which(names(fastaformat)==i))}
  fastaformat<-fastaformat[-id.remove]
  id.ref<-which(sapply(names(fastaformat),grepl,'EPI_ISL_402125')==T)
  mu<-mu_linear_reg_id(fastaformat,id.ref)
  samp_times<-samp_times[-id.remove]
  name_samp<-name_samp[-id.remove,]
  fastafile<-as.phyDat(fastaformat)
  tree<-serial_upgma(fastafile,mu, samp_times, name_samp, model="F81")
  plot(tree,show.tip.label = TRUE,cex=.3)
  print(paste("TMRCA is approximately at (note:this does not account correction of negative edges!!)", date_decimal(lastdate-tree$tmrca)))
}

```


Phylodynamic analysis and its plots.
```{r,echo=FALSE}
bnp<-BNPR(tree)
bnp_ps<-BNPR_PS(tree)

axlabs<-axis_label(bnp_ps,lastdate,byy=4/365)

plot_BNPR2(bnp,axlabs = axlabs)
plot_BNPR2(bnp_ps,axlabs = axlabs)
```

