---
title: "Phylodynamics using Serial UPGMA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the FASTA file of your county of interest. 
```{r, echo=FALSE}
library(phylodyn)
library(ape)
library(lubridate)
source("function_serial.R")
#Read fasta file
country<-"distFrance"
data<-paste(country,".RData",sep="")
load(data)
distList<-listout
```

distList should be a list such that
distList$n is the sample size
distList$seq_names should be the names of the sequence
distList$hamming is a R dist object of hamming matrices
distList$distGen is a R dist object of a genetic distance of choice

Extract sequences, sequence names, and sampling dates from the file
```{r, echo=FALSE}
#Save name of the sequence and sampling times from the file
n<-distList$n
seq_names<-distList$seq_names
samp_times<-c()
for (r in seq_names){
  samp_times<-c(samp_times,paste(strsplit(r,"|")[[1]][(length(strsplit(r,"|")[[1]])-9):length(strsplit(r,"|")[[1]])],collapse =""))
}
samp_times<-decimal_date(date(samp_times))
lastdate<-max(samp_times)
samp_times2<-max(samp_times)-samp_times
name_samp<-cbind(samp_times2,seq_names)
```

Ensure that there are at least two sequences at t=0, otherwise remove the sequence
```{r, echo=FALSE}
while(table(samp_times2)[1]==1){
  idx<-which(samp_times2==0)
  #Remove the sequence observed only once
  #fastaformat <- fastaformat[-idx]
  seq_names<-seq_names[-idx]
  samp_times2<-samp_times2[-idx]
  name_samp<-name_samp[-idx,]
  #Make the the last sequence the new 0
  #samp_times<-samp_times-min(samp_times)
  distList$distGen<-distList$distGen[-idx,-idx]
}
```

Compute mutation rate. Note: the reference sequence must be in the last spot
```{r,echo=FALSE}
mu<-mu_linear_reg_inputDist(distList)
```

Compute serial UPGMA tree and plot it
```{r,echo=FALSE}
tree<-serial_upgma_inputDist(distList,mu, samp_times2, name_samp)
plot(tree,show.tip.label = FALSE,cex=.3)
```




Phylodynamic analysis and its plots.
```{r,echo=FALSE}
bnp<-BNPR(tree)
bnp_ps<-BNPR_PS(tree)

axlabs<-axis_label(bnp_ps,lastdate,byy=4/365)

plot_BNPR2(bnp,axlabs = axlabs)
plot_BNPR2(bnp_ps,axlabs = axlabs)
```

