---
title: "Preliminary Molecular Clock Check and Summary Statistics"
output: pdf_document
---

```{r,echo=FALSE}
rm(list=ls())
library(phylodyn)
library(ape)
library(phangorn)
library(lubridate)

base.dir <- '~/Documents/Covid_Analysis/'
data.dir <- paste(base.dir, 'alignment/data/', sep='')

#ref.meta <- read.delim(paste(data.dir, 'genbank_MN908947_meta.txt',sep=''), header=T, as.is=T, sep='\t')

meta_fig <- read.delim2(paste(data.dir, '/all_meta.tsv', sep=''), quote="",
                           header=TRUE, sep='\t', as.is=T)

countr<-length(unique(meta_fig$country))
total.sam<-nrow(meta_fig)
main1<-paste("A total of ",total.sam," from ",countr, "countries --",date(),collapse="")
plot(sort(table(meta_fig$country)),cex.axis = 0.35,las=2,xlab="",ylab="Number of samples",main="")

```


```{r,echo=FALSE}
dates_samp<-ymd(meta_fig$date)
plot(table(dates_samp),cex.axis = 0.4,las=2,xlab="",ylab="Number of samples",main="Sequences by date")
setwd(data.dir)
system('curl https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv -o owid-covid-data.csv')

```


```{r,echo=FALSE}
root<-"Wuhan-Hu-1/2019"
ref<-which(meta_fig$strain==root)
##Sequence data (aligned already)
#this is where I used to compute the pairwise hamming distance that is difficult 
#will compute pairwise hamming in 4 batches
hamminglist<-0
fasta.out <- 'test.fasta'
source(paste(base.dir,"alignment/code/subset_data.R",sep=""))

for (j in seq(1,nrow(meta_fig),by=100)[-1]){
print(j)
subset.fasta(base.dir,c(ref,seq(j-100,j)), fasta.out)
gisaid.aligned <- paste(data.dir, 'test.fasta', sep='')
gisaidall <- read.FASTA(gisaid.aligned)
fastafile <- as.phyDat(gisaidall)
#Pairwise number of differences
hamminglist<-c(hamminglist,as.matrix(dist.hamming(fastafile,ratio=FALSE))[1,-1])
#plot(table(hamming)/2,xlab="Hamming Distance",ylab="Frequency")
}
subset.fasta(base.dir,c(ref,seq(j+1,length(meta_fig))), fasta.out)
gisaid.aligned <- paste(data.dir, 'test.fasta', sep='')
gisaidall <- read.FASTA(gisaid.aligned)
fastafile <- as.phyDat(gisaidall)
#Pairwise number of differences
hamminglist<-c(hamminglist,as.matrix(dist.hamming(fastafile,ratio=FALSE))[1,-1])

```


```{r,echo=FALSE}
plot(table(hamminglist),ylab="Frequency",xlab="Hamming distance to reference")
```

A rough estimate of the mutation rate can be obtained by regressing genetic divergence on time divergence. This estimate ignores correlation among samples, population structure and pairwise information. A more reliable estimate can be found in the mutation section.

```{r,echo=FALSE}
par(mfrow=c(1,1))
x<-dates_samp-dates_samp[ref]
reg<-lm(hamminglist~-1+x)
summary(reg)

plot(dates_samp-dates_samp[ref],hamming[ref,],ylab="Distance to reference",xlab="Time difference (Days)",main=paste("A total of ",round(reg$coefficients[[1]]*28,2)," of mutations per month",sep=""))
abline(reg)

library("ggplot2")
df <- data.frame(x_var = dates_samp-dates_samp[ref] 
             ,y_var = hamminglist
             )
ggplot(data = df, aes(x = x_var, y = y_var)) + theme_bw()+
  geom_point(color = 'red')+ggtitle(paste("A total of ",round(reg$coefficients[[1]]*28,2)," of mutations per month",sep=""))+
  stat_smooth(method = 'lm',formula='y~-1+x')+xlab("Time difference") + ylab("Distance to reference") +ylim(0,40)


tocheck<-which(hamminglist>quantile(hamminglist,.995))
tocheck
for (j in 1:length(tocheck)){
  text(x[tocheck[j]],hamminglist[tocheck[j]]+0.5,meta_fig[tocheck[j],1],cex=.5)
  print(meta_fig[tocheck[j],1])
}
```

To explore regions and number of polymorphic sites
```{r,echo=FALSE}
data.matrix<-as.character(as.matrix(fastafile))
ref.data<-data.matrix[ref,]
#subset only where ref has information
sum(ref.data!="-")
data.matrix<-data.matrix[,ref.data!="-"]
binary.matrix<-matrix(0,nrow=nrow(data.matrix),ncol=ncol(data.matrix))
divergence<-rep(0,ncol(data.matrix))
entropyl<-c(0,1)
library(entropy)
#Convert to binary matrix when possible
for (j in 1:ncol(data.matrix)){
  div<-table(data.matrix[,j])
  divergence[j]<-(sum(names(div)=="a")>0)+(sum(names(div)=="c")>0)+(sum(names(div)=="t")>0)+(sum(names(div)=="g")>0)
  if (divergence[j]>2){
    entropyl<-rbind(entropyl,c(j,entropy(div)))}
}

divsum<-table(divergence)
```

```{r,echo=FALSE}
orf1ab.range <- as.integer(ref.meta[ref.meta$name=='orf1ab', c('start', 'end')])
s.range <- as.integer(ref.meta[ref.meta$name=='S', c('start', 'end')])
e.range <- as.integer(ref.meta[ref.meta$name=='E', c('start', 'end')])
m.range <- as.integer(ref.meta[ref.meta$name=='M', c('start', 'end')])
n.range <- as.integer(ref.meta[ref.meta$name=='N', c('start', 'end')])

col.vec <- c("#E41A1C80", "#377EB880", "#4DAF4A80", "#984EA380", "#FF7F0080")
par(xpd=TRUE, mar=c(2,2,1,8))
plot(divergence, pch=19, col="blue", cex=0.7)
rect(orf1ab.range[1], 1 , orf1ab.range[2], 4,col=col.vec[1])
rect(s.range[1], 1 , s.range[2], 4,col=col.vec[2])
rect(e.range[1], 1 , e.range[2], 4,col=col.vec[3])
rect(m.range[1], 1 , m.range[2], 4,col=col.vec[4])
rect(n.range[1], 1 , n.range[2], 4,col=col.vec[5])
points(divergence,pch=19,col="blue", cex=0.7)
legend('right', legend=c('ORF1ab', 'S', 'E', 'M', 'N'), col=col.vec, bty='n',
       lwd=10, inset=c(-0.3,0))

x1<-max(entropyl[entropyl[,1]>.3,1][entropyl[entropyl[,1]>.3,1]<1000])
x2<-min(entropyl[entropyl[,1]>.3,1][entropyl[entropyl[,1]>.3,1]>25000])
```


Of the total `r ncol(data.matrix)` sites in the reference sequence, `r divsum[1]` are monomorphic sites, `r divsum[2]` are binary, and only `r divsum[3]+divsum[4]` have 3 or 4 types. Largest entropy is observed is in the first `r x1` bases and the last 26447 bases.



#Reference:

Rambaut, Andrew, Tommy T. Lam, Luiz Max Carvalho, and Oliver G. Pybus. "Exploring the temporal structure of heterochronous sequences using TempEst (formerly Path-O-Gen)." Virus evolution 2, no. 1 (2016).

