---
title: "Preliminary Molecular Clock Check and Summary Statistics"
output: pdf_document
---

```{r,echo=FALSE}
rm(list=ls())
library(phylodyn)
library(ape)
library(phangorn)
library(lubridate)

base.dir <- '~/Documents/Covid_Analysis/'
data.dir <- paste(base.dir, 'alignment/data/', sep='')

##Sequence data (aligned already)
gisaid.aligned <- paste(data.dir, 'aligned.fasta', sep='')
gisaidall <- read.FASTA(gisaid.aligned)

meta_fig <- read.delim(paste(data.dir, '/gisaid_meta_pp.tsv', sep=''), 
                           header=TRUE, sep='\t', as.is=T)


countr<-length(unique(meta_fig$country))
total.sam<-nrow(meta_fig)
main1<-paste("A total of ",total.sam," from ",countr, "countries --",date(),collapse="")
plot(sort(table(meta_fig$country)),cex.axis = 0.6,las=2,xlab="",ylab="Number of samples",main=main1)
```


```{r,echo=FALSE}
dates_samp<-ymd(meta_fig$date)
plot(table(dates_samp),cex.axis = 0.4,las=2,xlab="",ylab="Number of samples",main="Sequences by date")
```


```{r,echo=FALSE}
fastafile <- as.phyDat(gisaidall)
#Pairwise number of differences
hamming<-as.matrix(dist.hamming(fastafile))*as.numeric(summary(gisaidall)[1])

plot(table(hamming)/2,xlab="Hamming Distance",ylab="Frequency")
```


```{r,echo=FALSE}
root<-"Wuhan-Hu-1/2019"
ref<-which(meta_fig$strain==root)
plot(table(hamming[ref,]),ylab="Frequency",xlab="Hamming distance to reference")
```

A rough estimate of the mutation rate can be obtained by regressing genetic divergence on time divergence. This estimate ignores correlation among samples, population structure and pairwise information. A more reliable estimate can be found in the mutation section.

```{r,echo=FALSE}
par(mfrow=c(1,1))
x<-dates_samp-dates_samp[ref]
reg<-lm(hamming[ref,]~-1+x)
summary(reg)

plot(dates_samp-dates_samp[ref],hamming[ref,],ylab="Distance to reference",xlab="Time difference (Days)",main=paste("A total of ",round(reg$coefficients[[1]]*28,2)," of mutations per month",sep=""))
abline(reg)
tocheck<-which(hamming[ref,]>quantile(hamming[ref,],.995))
tocheck
for (j in 1:length(tocheck)){
  text(x[tocheck[j]],hamming[ref,tocheck[j]]+0.5,meta_fig[tocheck[j],1],cex=.5)
}
```

To explore regions and number of polymorphic sites
```{r,echo=FALSE}
data.matrix<-as.character(as.matrix(fastafile))
ref.data<-data.matrix[ref,]
#subset only where ref has information
sum(ref.data!="-")
data.matrix<-data.matrix[,ref.data!="-"]
binary.matrix<-matrix(0,nrow=nrow(data.matrix),ncol=ncol(data.matrix))
divergence<-rep(0,ncol(data.matrix))
entropyl<-c(0,1)
library(entropy)
#Convert to binary matrix when possible
for (j in 1:ncol(data.matrix)){
  div<-table(data.matrix[,j])
  divergence[j]<-(sum(names(div)=="a")>0)+(sum(names(div)=="c")>0)+(sum(names(div)=="t")>0)+(sum(names(div)=="g")>0)
  if (divergence[j]>2){
    entropyl<-rbind(entropyl,c(j,entropy(div)))}
}

divsum<-table(divergence)
plot(divergence,pch=19,col="blue",ylab="CNV")
abline(v=21563,lty=2)
abline(v=25384,lty=2)
x1<-max(entropyl[entropyl[,1]>.3,1][entropyl[entropyl[,1]>.3,1]<1000])
x2<-min(entropyl[entropyl[,1]>.3,1][entropyl[entropyl[,1]>.3,1]>25000])
```

Of the total `r ncol(data.matrix)` sites in the reference sequence, `r divsum[1]` are monomorphic sites, `r divsum[2]` are binary, and only `r divsum[3]+divsum[4]` have 3 or 4 types. Largest entropy is observed is in the first `r x1` bases and the last 26447 bases.



#Reference:

Rambaut, Andrew, Tommy T. Lam, Luiz Max Carvalho, and Oliver G. Pybus. "Exploring the temporal structure of heterochronous sequences using TempEst (formerly Path-O-Gen)." Virus evolution 2, no. 1 (2016).

