---
title: "Preliminary Molecular Clock Check and Summary Statistics"
output: pdf_document
---

```{r,echo=FALSE}
rm(list=ls())
library(phylodyn)
library(ape)
library(phangorn)
library(lubridate)

base.dir <- '~/Documents/Covid_Analysis/'
data.dir <- paste(base.dir, 'alignment/data/', sep='')

##Sequence data (aligned already)
gisaid.aligned <- paste(data.dir, 'aligned.fasta', sep='')
gisaidall <- read.FASTA(gisaid.aligned)

meta_fig <- read.delim(paste(data.dir, '/gisaid_meta_pp.tsv', sep=''), 
                           header=TRUE, sep='\t', as.is=T)


countr<-length(unique(meta_fig$country))
total.sam<-nrow(meta_fig)
main1<-paste("A total of ",total.sam," from ",countr, "countries --",date(),collapse="")
plot(sort(table(meta_fig$country)),cex.axis = 0.6,las=2,xlab="",ylab="Number of samples",main=main1)
```


```{r,echo=FALSE}
dates_samp<-ymd(meta_fig$date)
plot(table(dates_samp),cex.axis = 0.4,las=2,xlab="",ylab="Number of samples",main="Sequences by date")
```


```{r,echo=FALSE}
fastafile <- as.phyDat(gisaidall)
#Pairwise number of differences
hamming<-as.matrix(dist.hamming(fastafile))*as.numeric(summary(gisaidall)[1])

plot(table(hamming)/2,xlab="Hamming Distance",ylab="Frequency")
```


```{r,echo=FALSE}
root<-"Wuhan-Hu-1/2019"
ref<-which(meta_fig$strain==root)
plot(table(hamming[ref,]),ylab="Frequency",xlab="Hamming distance to reference")
```

A rought estimate of the mutation rate can be obtained by regressing genetic divergence on time divergence. This estimate ignores correlation among samples, population structure and pairwise information. A more reliable estimate can be found in the mutation section.

```{r,echo=FALSE}
par(mfrow=c(1,1))
x<-dates_samp-dates_samp[ref]
reg<-lm(hamming[ref,]~-1+x)
summary(reg)

plot(dates_samp-dates_samp[ref],hamming[ref,],ylab="Distance to reference",xlab="Time difference (Days)",main=paste("A total of ",round(reg$coefficients[[1]]*28,2)," of mutations per month",sep=""))
abline(reg)
tocheck<-which(hamming[ref,]>quantile(hamming[ref,],.995))
tocheck
for (j in 1:length(tocheck)){
  text(x[tocheck[j]],hamming[ref,tocheck[j]]+0.5,meta_fig[tocheck[j],1],cex=.5)
}
```


#Reference:

Rambaut, Andrew, Tommy T. Lam, Luiz Max Carvalho, and Oliver G. Pybus. "Exploring the temporal structure of heterochronous sequences using TempEst (formerly Path-O-Gen)." Virus evolution 2, no. 1 (2016).